### 2021-10-27 text/outline/outline.tex
ls

In this section we study the thresholding method from empirical and theoretical perspectives. First, we define some notation.

\blue{Let $n \approx 100,000 - 250,000$ be the number of cells in the experiment. Consider a single perturbation-gene pair. For cell $i \in \{ 1, \dots, n \},$ let $p_i \in \{ 0,1 \}$ indicate whether the cell was perturbed, $m_i \in \N$ be the number of observed gene UMIs, $g_i \in \N$ be the number of observed gRNA UMIs, $l^m_i \in \N$ be the gene library size, and $z_i \in \R^{d-1}$ be a vector of technical factors (e.g., batch, percent mitochondrial reads, etc.). The thresholding method is defined as follows:}
\blue{
\begin{itemize}
\item[1.] For given threshold $c \in \N$, calculate the imputed value $\hat{p}_i$ of $p_i$ by $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c. \end{cases}$$
\item[2.] Fit the regression model \cite{Sarkar2021}
$$ m_i | \left(\hat{p}_i, z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter, and $$\log\left(\mu_i\right) = \beta_m^0 + \beta_m \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
\item[3.] Fit a GLM to obtain an estimate $\hat{\beta}_m$ of $\beta_m$. Compute a confidence interval and $p$-value for $\beta_m$.
\end{itemize}}

@@

\blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}

\blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}

\blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets. There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}

\blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}

>>>>

In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several fundamental challenges of the method. Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$). Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced in the cell; let $g_i \in \N$ be the number of gRNA transcripts sequenced in the cell; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size) in the cell; and finally, let $z_i \in \R^{d-1}$ be the cell-specific vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.). The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
\begin{equation*}
\begin{cases}
m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i), \\
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
\end{cases}
\end{equation*}
where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term. Fit a GLM to obtain estimates of the parameters.

\item[3.] Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}

The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors. Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed. $\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively.

@@

We studied the behavior of the thresholding method on real data, uncovering an attenuation bias effect and a bias-variance tradeoff as a function of the selected threshold. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by setting the threshold to 1 were smaller in magnitude than those produced by setting the threshold to 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Fold change estimates generated by setting the threshold to 5 and setting the threshold to 20 were more concordant; however, the threshold of 20 yielded slightly stronger effect sizes (Figure \ref{thresholding_empirical}b). These results are consistent with an \textit{attenuation bias} phenomenon: small thresholds tend to underestimate the true parameter value, and this effect diminishes gradually as the threshold increases.

We hypothesized that selecting a large threshold comes with the cost of increasing the variance of the estimator.

% \blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}

% \blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}

% \blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets. There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}

% \blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}
